---
title: "QBS177_FINAL PROJECT"
author: "GROUP3"
date: "2/25/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
This data set contains 31 clinical attributes, m-RNA levels z-score for 331 genes, and mutation in 175 genes for 1904 breast cancer patients. Here, we just used the clinical variables of the data set. Clinical attributes in the dataset have 31 variables. 
```{r}
library(tidyverse)
```


```{r}
RNA_Mutation<-read.csv("METABRIC_RNA_Mutation.csv")#reading the data
```

```{r}
RNA_Mutation_updated<- RNA_Mutation[ -c(0,521:693) ] #removing all the mutation columns for better analysis purposes
```

```{r}
(is.na(RNA_Mutation_updated))#checking for NA
str(RNA_Mutation_updated)
```


```{r}
#To replace “ ” with  NA
RNA_Mutation_updated[RNA_Mutation_updated == ""] <- NA  
```

```{r}
# To change character variables to factor 
RNA_Mutation_updated$type_of_breast_surgery <- as.factor(RNA_Mutation_updated$type_of_breast_surgery)
RNA_Mutation_updated$cellularity <- as.factor(RNA_Mutation_updated$cellularity); 
RNA_Mutation_updated$chemotherapy <- as.factor(RNA_Mutation_updated$chemotherapy)
RNA_Mutation_updated$pam50_._claudin.low_subtype <- as.factor(RNA_Mutation_updated$pam50_._claudin.low_subtype); 
RNA_Mutation_updated$er_status <- as.factor(RNA_Mutation_updated$er_status); 
RNA_Mutation_updated$neoplasm_histologic_grade <- as.factor(RNA_Mutation_updated$neoplasm_histologic_grade);
RNA_Mutation_updated$her2_status <- as.factor(RNA_Mutation_updated$her2_status);
RNA_Mutation_updated$tumor_other_histologic_subtype <- as.factor(RNA_Mutation_updated$tumor_other_histologic_subtype)
RNA_Mutation_updated$hormone_therapy <- as.factor(RNA_Mutation_updated$hormone_therapy);
RNA_Mutation_updated$inferred_menopausal_state<- as.factor(RNA_Mutation_updated$inferred_menopausal_state)
RNA_Mutation_updated$integrative_cluster<- as.factor(RNA_Mutation_updated$integrative_cluster ); 
RNA_Mutation_updated$primary_tumor_laterality<- as.factor(RNA_Mutation_updated$primary_tumor_laterality)
RNA_Mutation_updated$pr_status<- as.factor(RNA_Mutation_updated$pr_status); 
RNA_Mutation_updated$radio_therapy <- as.factor(RNA_Mutation_updated$radio_therapy)
RNA_Mutation_updated$overall_survival<- as.factor(RNA_Mutation_updated$overall_survival); 

```

```{r}
# exploring numeric data
# To remove Redundant  column
RNA_Mutation_updated$patient_id <- NULL        
RNA_Mutation_updated$cancer_type <- NULL       
RNA_Mutation_updated$death_from_cancer <- NULL  
RNA_Mutation_updated$cancer_type_detailed <- NULL  
RNA_Mutation_updated$oncotree_code <- NULL   
RNA_Mutation_updated$cohort <- NULL   
RNA_Mutation_updated$tumor_stage <- NULL
RNA_Mutation_updated$overall_survival_months  <- NULL
RNA_Mutation_updated$X3.gene_classifier_subtype  <- NULL
RNA_Mutation_updated$er_status_measured_by_ihc  <- NULL
RNA_Mutation_updated$her2_status_measured_by_snp6  <- NULL


clinical<-RNA_Mutation_updated[, c(1:20)]    # To make clinical data set
```


```{r}
clinical_n <- select_if(clinical, is.numeric) 
corrmatrix <- cor(clinical_n)

clinical <- na.omit(clinical)

corrmatrix <- cor(clinical_n)

corrplot::corrplot(corrmatrix, method="shade", type = "upper",tl.cex=.6	
, tl.col="black", title="Correlation Plot",number.font = 2, mar=c(0,0,1,0),	)
```
```{r}
library(ggformula)
clinical %>% gf_boxplot(age_at_diagnosis~overall_survival) %>%   
  gf_labs(x = "Survival",  
    y = "Age") %>%   
  gf_theme(axis.text = element_text(size = 12),
           axis.title = element_text(size = 12),
        legend.title=element_text(size=14), 
    legend.text=element_text(size=9))


clinical %>% gf_boxplot(tumor_size~overall_survival) %>%   
  gf_labs(x = "Survival",  
    y = "Tumor size") %>%   
  gf_theme(axis.text = element_text(size = 12),
           axis.title = element_text(size = 12),
        legend.title=element_text(size=14), 
    legend.text=element_text(size=9))

```


```{r}
data.nona=RNA_Mutation_updated[complete.cases(RNA_Mutation_updated),] #delete all rows with any NA
genes=data.nona[,21:509] #subset gene columns
```


```{r}
#PCA analysis
pca=prcomp(genes,scale=TRUE)
summary(pca)
```

```{r}
#install.packages("factoextra")
library(factoextra)
fviz_pca_ind(pca, geom.ind = "point", pointshape = 21, 
             pointsize = 2, 
             fill.ind = data.nona$overall_survival, 
             col.ind = "black", 
             palette = "jco", 
             addEllipses = TRUE,
             label = "var",
             col.var = "black",
             repel = TRUE,
             legend.title = "Diagnosis") +
  ggtitle("2D PCA-plot from 30 feature dataset") +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
library(ggplot2)
screeplot(pca, type = "l", npcs = 100, main = "Screeplot of the first 100 PCs")
abline(h = 1, col="red", lty=5)
legend("topright", legend=c("Eigenvalue = 1"),
       col=c("red"), lty=5, cex=0.6)
cumpro <- cumsum(pca$sdev^2/sum(pca$sdev^2))
plot(cumpro[0:200], xlab = "PC #", ylab = "Amount of explained variance", main = "Cumulative variance plot")
abline(v = 100, col="blue", lty=5)
abline(h = 0.7, col="blue", lty=5)
legend("topleft", legend=c("Cut-off @ PC100"),
       col=c("blue"), lty=5, cex=0.6)
```

```{r}
comp=data.frame(pca$x[,1:100]) #retain the first 100 principal components
cluster.data=cbind(data.nona[,1:20],comp) #dataset for cluster analysis
```

total month of surviving - linear regression 

```{r}
data.yoo<-read.csv("METABRIC_RNA_Mutation.csv")
model1_linear<-lm(overall_survival_months~age_at_diagnosis+lymph_nodes_examined_positive+mutation_count+nottingham_prognostic_index+tumor_size,data=data.yoo)
summary(model1_linear)
```

```{r}
model2_linear<-lm(charges~age+bmi+smoker,data=insurance)
summary(model2_linear)
```

```{r}
insurance$charges_cat <- ifelse(insurance$charges >median(insurance$charges), 1, 0)
```

```{r}
model_glm<-glm(charges_cat~age+sex+bmi+children+smoker+region,family=binomial,data=insurance)
summary(model_glm)
```

K-Means Cluster Analysis

```{r}
#looking purely at gene expression subset of the data - numeric values for kmeans clustering
kvars <- paste("PC", 1:100, sep = "")
k.df <- cluster.data[kvars]
head(k.df)
```
```{r}
#scale the data - no dependency to arbitrary variable unit
k.df <- scale(k.df)
head(k.df)
```
```{r}
#euclidean distance
k.dist <- get_dist(k.df)
k.dist
```
```{r}
k.check <- kmeans(k.df, centers = 2, nstart = 25)
k.check
```
Two clusters gives clusters with sizes 205 and 1403

```{r}
fviz_cluster(k.check, data = k.df)
```
Starting with two clusters, we can see there is serious overlap in clusters. There is no 
clear division. We can find optimal cluster number with log-elbow method

```{r}
#compute total within cluster sum of squares
set.seed(429)
wss <- function(k) {
  kmeans(k.df, k, nstart = 10 )$tot.withinss
}

#consider 1 to 100 clusters
kc.vals <- 1:150

#extract within cluster sum of squares for all number of clusters
wss.vals <- map_dbl(kc.vals, wss)

#plot log-elbow method
plot(kc.vals, wss.vals,
       type="b", pch = 19, frame = FALSE, 
       xlab="Number of clusters K",
       ylab="Total within-clusters sum of squares")
```

We can say about 20 clusters looking at this. We can double check as well. From here we can 
move onto the optimal model

```{r}
set.seed(429)

fviz_nbclust(k.df, kmeans, method = "wss")
```


```{r}
set.seed(429)
k.fin <- kmeans(k.df, 20, nstart = 25)
k.fin
```
```{r}
fviz_cluster(k.fin, data = k.df)
```
We can see that K-Means clustering did yield clusters of genes that have similar properties 
and while the plots do look chaotic, we can attribute this to the number of dimensions 
associated with the data. From the PCA, we yielded about 100 genes to consider, and we use 
them all in our clustering analysis. 














