---
title: "QBS121_Final_Presentation_AvaniKuthe"
author: "Avani Kuthe"
date: "3/3/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Summary of LMM and GEE Analysis

The data is about individuals with breast cancer, meaning that certain gene expressions will 
be similar, and repreated within the data set, as well as tumor measurements. Patients can be
clustered by Tumor diagnosis, either benign and malignant. Because of this, we will be 
looking into LMM and GEE. Firstly, we'll look at correlation and discern what correlation 
structure the data falls under, and we will also decide to treat clusters as fixed effects. 
Family is binomial. 
We use a linear mixed model to take into account that tumor measurements are fixed 
effects, and additionally we might not be able to assume individual residuals between 
variables, meaning that the variables may not be independent. Because of that, there is 
possible confounding, in that each attribute could affect the next (ie. radius could affect 
perimeter, or area, and so on). I created LMM model looking at the means of each attribute 
(Taken a subset of the overall data) and employed the Restricted Maximum Likelihood method, 
which takes into account any possible bias in standard deviations, which we’ve measure per 
observation in the dataset already, so this works to remove biases that come from those 
standard deviation measurements. With this output we can see that Radius, Texture, Perimeter,
Area, Smoothness, Concavity, and Symmetry affect tumor diagnosis (B or M) most significantly.
In the previous LMM model, we did not take correlation into account, so we can expand the LMM
model with correlation analysis included. I create a correlation plot and found that it most 
closely resembles an autoregressive correlation structure, though we do have to acknowledge it is not perfectly autoregressive. Next we create our GEE model, using the means of all the 
tumor attributes, as well as specifying the family as binomial, because of B or M target 
variable, which I did turn to a numeric either 1 for M or 0 for B. We can see that the list of 
most statistically significant predictors narrows to give us Texture, Area, Smoothness, and 
Concavity. Overall, we have taken into account possible confounding and clustering affects 
present in the data to give a model that is robust enough to predict the most statistically 
significant tumor attributes to diagnose whether a tumor is B or M

## Prelinimary data wrangling

```{r}
library(readr)
breast_cancer <- read_csv("breast-cancer.csv")
#View(breast_cancer)
```
```{r}
#make a numerical binary value for the target data variable 
dat1 <- breast_cancer
dat1$diagnosis <- ifelse(dat1$diagnosis=="M",1,0)
#head(dat1)
```

## Correlation Plot

```{r}
#correlation plot to determine corrleation structure
library(corrplot)
library(RColorBrewer)
M <-cor(dat1)
corrplot(M, type="upper", order="hclust",
         col=brewer.pal(n=8, name="RdYlBu"))
```
Looking at this correlation plot, we can assume that the correlation structure is 
autoregressive. Now with this, we can fit our linear mixed model

## LMM

```{r}
library(lme4)
library(nlme)
```

```{r}
#LMM Model
#for the purposes of our analysis we will look at a subset of the data that considered the mean values of tumor measures
#looking at lmm because we first want to consider tumor values as fixed affects, under the assumption that certain gene expression that come with malignant or benign diagnosis are the same 
dat.mn.vals <- dat1[c(1:12)]
#head(dat.mn.vals)
mod1 <- lme(fixed = diagnosis~radius_mean+texture_mean+perimeter_mean+area_mean+smoothness_mean+compactness_mean+concavity_mean+symmetry_mean+fractal_dimension_mean, random = ~1|id, data = dat.mn.vals)
summary(mod1)
```
we can see here that radius, texture, perimeter, area, smoothness, concavity, and symmetry 
are significant in detecting diagnosis type for breast cancer. Additionally, we can see that 
compactness and fractal dimension are not significant.

## GEE Model

```{r}
#GEE Model
library(geepack)
```

Extend the LMM model with GEE to allow for the correlation of the variables

```{r}
#GEE Model
mod2 <- geeglm(diagnosis~radius_mean+texture_mean+perimeter_mean+area_mean+smoothness_mean+compactness_mean+concavity_mean+symmetry_mean+fractal_dimension_mean, family = binomial, data = dat.mn.vals, id = id, corstr = "ar1")
summary(mod2)
```

So here we can see with the specification of the correlation structure, we can narrow down 
our list of significant tumor attributes that affect breast cancer diagnosis. In this case, 
we can see that texture, area, smoothness, and concavity are the most significant predictors.






